---
title: "Imputation Benchmarking on Nottingham Temperature Data"
author: "blesson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(42)
```

## 1. Introduction

This report demonstrates the `imputeTestbench` package on the `nottem`
dataset — monthly mean air temperatures (in Fahrenheit) recorded at
Nottingham Castle, England, from 1920 to 1939. It contains 240 monthly
observations with no missing values, making it a clean baseline for
controlled imputation benchmarking.

Temperature series like this are common in environmental and climate
research, where sensor failures and data transmission errors routinely
cause missing readings. Understanding which imputation method best
recovers the original values is practically important.

```{r load}
library(imputeTestbench)

temp_data <- as.numeric(nottem)

cat("Observations:", length(temp_data), "\n")
cat("Period      : 1920 to 1939 (monthly)\n")
cat("Range       :", round(min(temp_data), 1), "to",
    round(max(temp_data), 1), "F\n")
cat("Mean        :", round(mean(temp_data), 1), "F\n")
```

---

## 2. Exploring the data

```{r eda, fig.width=10, fig.height=4}
plot(nottem,
     main = "Nottingham Castle: monthly mean temperature (1920-1939)",
     ylab = "Temperature (F)",
     xlab = "Year",
     col  = "steelblue",
     lwd  = 1.5)
abline(h   = mean(temp_data),
       col = "tomato",
       lty = 2,
       lwd = 1.2)
legend("topright",
       legend = paste0("Series mean (", round(mean(temp_data), 1), " F)"),
       col = "tomato", lty = 2, bty = "n")
```

The series shows a clear and consistent seasonal cycle — cold winters
around 35F, warm summers around 65F — repeating every 12 months across
20 years. This regularity is exactly what separates good imputation
methods from poor ones: a method that understands seasonality will
recover missing winter values as cold, not as the annual mean of ~49F.

---

## 3. Running the benchmark

We test four methods that represent increasing levels of sophistication:

- `na.mean` — replaces every missing value with the series mean
- `na.locf` — carries the last observed value forward
- `na.approx` — linear interpolation between surrounding values
- `na.interp` — seasonal decomposition-aware interpolation

```{r benchmark}
methods_to_test <- c("na.mean", "na.locf", "na.approx", "na.interp")

results <- impute_errors(
  dataIn          = temp_data,
  smps            = "mcar",
  methods         = methods_to_test,
  missPercentFrom = 10,
  missPercentTo   = 50,
  interval        = 10
)

print(results)
```

---

## 4. Visualising the results

```{r plot-results, fig.width=9, fig.height=5}
plot_errors(results, plotType = "line")
```

`na.mean` is consistently the worst performer. This makes intuitive sense:
the series mean is ~49F, but the actual values span 35F to 65F in a
predictable seasonal pattern. Filling a missing July reading with 49F
introduces a large error when the true value is closer to 63F.

`na.interp` performs best across all missingness levels because it
decomposes the series into trend and seasonal components before
interpolating, allowing it to "know" that a missing summer month
should be warm.

---

## 5. Seeing the imputation at 40% missingness

```{r impute-plot, fig.width=10, fig.height=7}
plot_impute(dataIn = temp_data, missPercent = 40)
```

At 40% missingness the differences become visually obvious. The flat
pink line from `na.mean` completely ignores the seasonal cycle.
`na.interp` produces imputed points that follow the seasonal wave
closely, making them nearly indistinguishable from the real observations.

---

## 6. How much better is the best method?

```{r quantify}
rmse_at_30 <- data.frame(
  method = methods_to_test,
  rmse   = round(c(results$na.mean[3],
                   results$na.locf[3],
                   results$na.approx[3],
                   results$na.interp[3]), 3)
)
rmse_at_30 <- rmse_at_30[order(rmse_at_30$rmse), ]
rmse_at_30$times_worse <- round(rmse_at_30$rmse / min(rmse_at_30$rmse), 2)
print(rmse_at_30)
```

```{r bar, fig.width=7, fig.height=5}
colors <- c("na.interp" = "steelblue", "na.approx" = "steelblue4",
            "na.locf"   = "orange",    "na.mean"   = "tomato")

bp <- barplot(rmse_at_30$rmse,
              names.arg = rmse_at_30$method,
              col       = colors[rmse_at_30$method],
              main      = "RMSE at 30% missingness — lower is better",
              ylab      = "RMSE",
              ylim      = c(0, max(rmse_at_30$rmse) * 1.15),
              las       = 2)
text(bp, rmse_at_30$rmse + 0.15,
     labels = rmse_at_30$rmse, cex = 0.85)
```

At 30% missingness, `na.mean` produces an error roughly
`r rmse_at_30$times_worse[rmse_at_30$method == "na.mean"]`x larger
than `na.interp`. The difference grows as missingness increases.

---

## 7. Conclusion

For a strongly seasonal series like Nottingham temperatures,
`na.interp` is clearly the best choice. The key takeaway is that
**method selection should match the structure of the series**: a method
that can represent seasonality will always outperform one that ignores
it, and the gap widens as the proportion of missing data grows.

`imputeTestbench` makes this comparison reproducible and systematic —
running a rigorous multi-method benchmark takes just a few lines of code.

---

## 8. Session info

```{r session}
sessionInfo()
```
