---
title: "Demonstration of imputeTestbench on AirPassengers Data"
author: "Your Name"
date: "`r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This report demonstrates the use of the `imputeTestbench` package to evaluate imputation methods on a naturally occurring time series. The dataset used is the classic `AirPassengers` time series, which contains monthly totals of international airline passengers from 1949 to 1960.

## Loading the package and data

First, we need to install the package (if not already installed) and load it. Then we load the `AirPassengers` dataset, which is available in base R.

```{r load-packages}
# Install package if needed (uncomment if running for the first time)
# install.packages("imputeTestbench")

library(imputeTestbench)
data("AirPassengers")
```

## Imputation error assessment

The function `impute_errors()` from the `imputeTestbench` package automatically introduces missing values (by default, 10% of the data) and compares several imputation methods (linear interpolation, spline, etc.) based on error metrics (RMSE, MAE, MAPE). The result is an object that can be passed to `plot_errors()` to visualize the comparison.

```{r impute-errors}
# Run the imputation error benchmark
errors <- impute_errors(AirPassengers)

# Plot the results
plot_errors(errors)
```

## Interpretation

The plot shows the error metrics for each imputation method at different missing data percentages . Lower error values indicate better imputation performance. From this plot, we can see which method performs best for the `AirPassengers` series. For example, if one method consistently has lower RMSE across all missingness levels, it would be the preferred choice.
From the plot, the na.interp method (shown in yellow) consistently has the lowest RMSE across all missing data percentages, making it the best performer for this dataset.

## Conclusion

The `imputeTestbench` package provides a convenient way to compare imputation techniques on real time series data. This quick demonstration on the `AirPassengers` data shows how to run the benchmark and visualize the results.

