---
title: "Structured Missing Data Simulation for Multivariate Time Series"
subtitle: "Extending imputeTestbench for Multivariate Support"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    number_sections: true
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = FALSE,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

---

# Why Multivariate?

`imputeTestbench` currently excels at benchmarking imputation methods on **univariate** time series. But real data is rarely univariate. Weather stations record temperature, humidity, and pressure simultaneously; financial datasets include multiple correlated indicators.

To evaluate imputation algorithms realistically, we need to:

- Simulate missingness that can affect several variables at once (e.g., a sensor array failure)
- Test methods that use cross-variable relationships (e.g., MICE, VAR imputation, `mtsdi`)
- Assess performance **per variable** — which sensors are hardest to impute?

Adding multivariate support will make `imputeTestbench` useful for a much wider range of applications.

---

# Current Univariate Architecture

The existing pipeline for a single time series (vector or `ts`) is:

1. **`sample_dat()`** — takes a complete series and returns a list of *repetition* corrupted copies, each with exactly a given percentage `b` of `NA`s (MCAR or block MAR).

2. **`impute_errors()`** — loops over:
   - Missing percentages (e.g., 10%, 20%, …)
   - Imputation methods (e.g., `na.approx`, `na.interp`)
   - Repetitions (each corrupted copy from `sample_dat`)
   
   For each repetition:
   - Apply the method to fill `NA`s → `filled`
   - Compute an error metric (RMSE, MAE, …) between original and `filled`
   - Store the single error value

**Storage:** Raw errors are kept in `errall[[method]][[percentage_index]]` as a vector of length `repetition`.

**Final output:** An `errprof` object — a list with `Parameter`, `MissingPercent`, and for each method a vector of average errors.

Everything is designed for vectors. To support matrices (rows = time, columns = variables), we must extend it without breaking existing code.

---

# Proposed Multivariate Extensions

We will **extend** the existing functions — not create separate `*_mv` versions. This keeps the API clean and uses all current features (custom methods, extra arguments, user-defined error functions).

## Input Detection

At the top of both `sample_dat` and `impute_errors`:

```{r input-detection}
if (is.matrix(dataIn) || is.data.frame(dataIn)) {
  multivariate <- TRUE
  dataIn <- as.matrix(dataIn)
  n_vars <- ncol(dataIn)
  if (n_vars == 1) multivariate <- FALSE   # treat single column as univariate
} else {
  multivariate <- FALSE
}
```

All subsequent logic branches on `multivariate`.

## Missing Data Generation — Extending `sample_dat()`

For multivariate input, `sample_dat` will return a list of corrupted **matrices** (one per repetition). New arguments control the missingness mechanisms:

| Argument | Description |
|---|---|
| `mv_mechanism` | One of `"mcar"`, `"mar"`, `"block"`, `"system_blackout"`. Default `"mcar"`. |
| `mv_mar_func` | A function that takes the full data matrix and returns a probability matrix of the same dimensions (for flexible MAR). |
| `mv_block_size` | Block length (can be a scalar or a vector of length `n_vars`). |
| `mv_n_blocks` | Number of blocks per variable (for independent block missingness). |
| `mv_synchronize` | Logical; if `TRUE`, blocks occur at the same rows for all variables (simulates a system-wide outage). |
| `mv_protect_ends` | Logical (default `TRUE`) to keep first and last time points observed, matching the univariate behaviour. |

### How It Works Internally

For each repetition, we generate a logical **mask** — a matrix of the same dimensions as the data, with `TRUE` where we want `NA`. Then we apply it:

```{r mask-apply}
corrupted <- data
corrupted[mask] <- NA
```

The mask is created by helper functions specific to each mechanism:

```{r mask-mechanisms}
# MCAR: each cell independently missing with probability b/100
mask <- matrix(runif(n * n_vars) < b / 100, nrow = n, ncol = n_vars)

# System blackout: all variables missing for same contiguous block
mask <- matrix(FALSE, n, n_vars)
start <- sample(2:(n - block_size), 1)   # protect ends
mask[start:(start + block_size - 1), ] <- TRUE

# Independent blocks: each variable gets its own blocks
mask <- matrix(FALSE, n, n_vars)
for (j in 1:n_vars) {
  for (k in 1:n_blocks[j]) {
    start <- sample(1:(n - block_size[j]), 1)
    mask[start:(start + block_size[j] - 1), j] <- TRUE
  }
}

# MAR: user function provides probabilities
prob_matrix <- mv_mar_func(data)        # user-defined
mask <- matrix(runif(n * n_vars) < prob_matrix, nrow = n, ncol = n_vars)
```

### Extended `sample_dat()` Skeleton

```{r sample-dat-skeleton}
sample_dat <- function(datin, ..., mv_mechanism = "mcar",
                       mv_mar_func = NULL,
                       mv_block_size = 10,
                       mv_n_blocks = 1,
                       mv_synchronize = FALSE,
                       mv_protect_ends = TRUE) {

  if (!multivariate) {
    # existing univariate code ...

  } else {
    out <- vector("list", repetition)

    for (i in 1:repetition) {
      mask <- switch(mv_mechanism,
        mcar = matrix(runif(n * n_vars) < b / 100, n, n_vars),
        block = generate_block_mask(
          n, n_vars, mv_block_size, mv_n_blocks,
          mv_synchronize, mv_protect_ends
        ),
        system_blackout = generate_system_blackout_mask(
          n, n_vars, mv_block_size, mv_protect_ends
        ),
        mar = generate_mar_mask(datin, mv_mar_func, ...)
      )

      corrupted <- datin
      corrupted[mask] <- NA
      out[[i]] <- corrupted
    }

    return(out)
  }
}
```

The helper functions (`generate_block_mask`, etc.) are implemented in separate internal files with unit tests.

## Integration with `impute_errors()`

The three nested loops remain unchanged. The critical modifications are inside the innermost repetition loop and in the error storage.

### Per-Variable Error Computation

```{r per-variable-error}
errs <- lapply(out, function(y) {   # y is a corrupted matrix
  filled <- eval(parse(text = toeval))  # imputed matrix (same dimensions)

  if (multivariate) {
    # Compute error per variable using existing vector error functions
    err_vec <- numeric(n_vars)
    for (j in 1:n_vars) {
      err_vec[j] <- eval(
        parse(text = paste0(errorParameter, "(dataIn[,j], filled[,j])"))
      )
    }
    return(err_vec)   # vector of length n_vars

  } else {
    # univariate: single error value
    return(eval(parse(text = paste0(errorParameter, "(dataIn, filled)"))))
  }
})
```

### Matrix Storage

After processing all repetitions for a given method and percentage, we combine the list of error vectors into a matrix:

```{r matrix-storage}
err_matrix <- do.call(rbind, errs)   # dimensions: repetition x n_vars
errall[[method]][[x]] <- err_matrix
```

Each cell of `errall` now stores a **matrix** instead of a vector.

## Final Output Structure

The final summary (the returned `errprof` object) is built by averaging over repetitions per variable:

```{r final-output}
out <- lapply(errall, function(method_list) {
  avg_per_perc <- lapply(method_list, colMeans)   # list of vectors
  do.call(rbind, avg_per_perc)                    # matrix: percentages x variables
})

names(out) <- methods
out <- c(list(Parameter = errorParameter, MissingPercent = percs), out)
class(out) <- c("errprof", "list")
attr(out, "errall") <- errall
```

For each method, the user now gets a **matrix** where:

- **Rows** = missing percentages
- **Columns** = variables

This immediately shows how a method performs on each variable separately.

## Backward Compatibility

- All changes are conditional on input type. If `dataIn` is a vector or `ts`, `multivariate = FALSE` and the original univariate code path is used.
- New `mv_*` arguments are simply ignored for univariate input, so existing scripts continue to work unchanged.
- Existing unit tests for univariate data must pass without modification.

---

# Performance Optimisation

Multivariate benchmarking can be computationally heavy. We will add an optional `parallel` argument to `impute_errors()` (default `FALSE`). When `TRUE`, we use the `future` and `future.apply` packages to parallelise the outer loop over percentages. This is safe because percentages are independent.



---


---

# Why This Design Is Better

| Feature | Benefit |
|---|---|
| **Seamless integration** | Users don't need to learn new functions |
| **Handles repetitions correctly** | Returns a list of corrupted matrices, matching the univariate contract |
| **Per-variable error storage** | Matrices inside `errall` are a natural extension of vectors |
| **Realistic missingness** | Multiple blocks, system blackouts, flexible MAR |
| **Performance built-in** | Optional parallelisation directly in the loop |
| **Backward compatible** | Existing univariate code unchanged |

---

# Future Extensions: True Multivariate Imputation Methods

The current proposal focuses on the infrastructure for multivariate missing data **simulation and evaluation**. The imputation methods themselves (e.g., `na.approx`, `na.interp`) are still applied column-wise and do not exploit correlations between variables.

However, the framework we build will be ready to accommodate truly multivariate methods that use cross-variable relationships, such as:

- **`mtsdi`** — EM-based imputation for multivariate time series
- **`mice`** with time series adaptations
- **`Amelia`** — bootstrapping + EM for time-series cross-sectional data
- **`missMDA`** — PCA-based imputation
- **VAR models** — vector autoregressive imputation

---


